---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---
```{r load libraries, echo=FALSE, include=FALSE}
library(dplyr)
library(lubridate)
library(readr)
library(ggplot2)
library(knitr)
```


## Loading and preprocessing the data
```{r load_data, echo=TRUE}
activity <- read.csv("activity.csv", stringsAsFactors = FALSE)
# convert date string to date using lubridate
activity$date <- ymd(activity$date)

```



## What is mean total number of steps taken per day?
```{r mean_per_day, echo = TRUE}
activity <- group_by(activity, date)
steps_perday <- aggregate(activity[c("steps", "interval")], by = list(as.factor(activity$date)), sum)

# create histogram
ggplot(steps_perday, aes(steps)) + geom_histogram() + xlab("steps per day")

mean_perday <- round(mean(steps_perday$steps, na.rm = TRUE), 2)
median_perday <- median(steps_perday$steps, na.rm = TRUE)

activity <- as.data.frame(ungroup(activity))
```
The mean per day is `r as.character(mean_perday)` and the median is `r median_perday`.

## What is the average daily activity pattern?
```{r daily_activity_pattern, echo=TRUE}
activity <- group_by(activity, interval)
steps_perinterval <- aggregate(activity[c("steps")], by = list(interval = activity$interval), mean, na.rm = TRUE)

# create time series
timeseries = format( seq.POSIXt(as.POSIXct(Sys.Date()), as.POSIXct(Sys.Date()+1), by = "5 min"), "%H%M", tz="GMT")
timeseries <- parse_time(timeseries[1:length(timeseries)-1], "%H%M")
steps_perinterval$timeseries <- timeseries

ggplot(steps_perinterval, aes(timeseries, steps)) + geom_line()

# find interval with max average steps
idx <- which.max(steps_perinterval$steps)
maxinterval <- steps_perinterval$timeseries[idx]

activity <- ungroup(activity)
```
The most steps occur on average in the 5 minutes after `r as.character(maxinterval)`.


## Imputing missing values
```{r missing_values, echo = TRUE}
na <- sum(is.na(activity$steps))
```
Of the `r nrow(activity)` step entries `r na` are missing, that is `r round(mean(is.na(activity$steps))*100, 2)`%.  

The missing values in the raw data will be filled with the mean number of steps for the corresponding time interval.
```{r fill_nas, echo=TRUE}
activity_filled <- activity
for (i in 1:nrow(activity)) {
    if (is.na(activity$steps[i])==TRUE){
        activity_filled$steps[i] <- steps_perinterval[(steps_perinterval$interval == activity$interval[i]),]$steps
    }
}

activity_filled <- group_by(activity_filled, date)
steps_perday_filled <- aggregate(activity_filled[c("steps")], by = list(as.factor(activity_filled$date)), sum)

# create histogram
ggplot(steps_perday_filled, aes(steps)) + geom_histogram() + xlab("steps per day")

mean_perday_filled <- round(mean(steps_perday_filled$steps, na.rm = TRUE), 2)
median_perday_filled <- median(steps_perday_filled$steps, na.rm = TRUE)

comp <- data.frame("mean" = c(mean_perday, mean_perday_filled), "median" = c(median_perday, median_perday_filled), row.names = c("with na", "imputed na"))
```

```{r table_output, echo = TRUE, results = "asis"}
kable(comp)
```
  
The results dont have an effect on the mean per day, and only a minor effect on the median.  
The total number of steps per day is affected because the imputed valus are added to the number of total steps.

## Are there differences in activity patterns between weekdays and weekends?
```{r day_of_the_week, echo = TRUE}
activity <- mutate(activity, "weekday" = weekdays(activity$date, 1))

wd_we <- c()
for (i in 1:nrow(activity)){
    if (activity$weekday[i] %in% c("Sa", "So")){
        wd_we[i] <- "weekend"
    } else {
        wd_we[i] <- "weekday"
    }
}

activity <- mutate(activity, "we_wd" = as.factor(wd_we))

ggplot(activity, aes(interval, steps)) + geom_line() + facet_grid(we_wd ~ .)
```

